---
title: 'SHARP Multiomics Workshop: Advanced Interactions Lab'
author: "Natalia Zemlianskaia"
date: "`r format(Sys.time(), '%d %B %Y')`"

output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    df_print: paged
---

```{r overall setup, include=FALSE, echo=FALSE}
library(gesso)

options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE)

# suppress warning messages for final rendering
old.warn <- getOption("warn")
options(warn=-1)

```



# Data exploration

Our research question is to discover important genotype by maternal smoking interactions that are associated with a child’s BMI at 6-11 years of age

The sample size is 1,122 children.
The outcome in our analysis is a child's BMI with 2 categories, 0: Thinness and Normal, 1: Overweight and Obese. Genotype matrix includes 1000 SNPs with additive allele coding (0/1/2). 

Exposure variable is binarized maternal active tobacco smoke during pregnancy, 0: no smoking, 1: smoking.

Adjustment variables include maternal pre-pregnancy BMI, age, education status, cohort inclusion, and child's sex and ethnicity (PCs).

For interaction analysis we have 1000 SNPs + 1 E + 1000 [SNPs by E] terms + 12 adjustment variables = 2013 predictors in total. 
```{r}
setwd("/cloud/project/Data")

# OUTCOME  
# hs_bmi_c_cat: body mass index categories at 6-11 years old - 
# WHO reference (1: Thinness, 2: Normal, 3:Overweight, 4: Obese)
# Y <- ifelse(as.numeric(Y) >= 3, 1, 0)
Y = readRDS("Y.rds")
sum(Y)/length(Y)

# EXPOSURE  
# e3_asmokcigd_p_None: maternal active tobacco smoke during pregnancy, mean num cig/day
# E <- ifelse(E > 0, 1, 0)
E = readRDS("E.rds")
sum(E)/length(E)

# GENOTYPES additive dosages for SNPs
X = readRDS("X.rds")
dim(X); n = dim(X)[1]; p = dim(X)[2]
head(X[,1:5])
MAF = colSums(X)/(2 * n)
summary(MAF)

# ADJUSTMENT VARIABLES
U = readRDS("U.rds")
colnames(U)
# h_mbmi_None: maternal pre-pregnancy body mass index
# e3_sex_Nonefemale: child sex (female / male)
# h_age_None: maternal age (years)
# h_cohort: cohort of inclusion (1 to 6) maternal
# h_edumc_None: maternal education (1: primary school, 2:secondary school, 
#               3:university degree or higher)
# ethn_PC1, ethn_PC2: PCs

causalSNPs = as.vector(unlist(read.table("CausalSNPs.txt")))
causalSNPs
```

# Analysis with gesso

`gesso` method focuses on a single environmental exposure and induces a main-effect-before-interaction hierarchical structure for the joint selection of interaction terms in a regularized regression model.
```{r}
# gesso (G[by]E[la]sso) analysis - hierarchical joint selection of GxE
start = Sys.time()
gesso_cv = gesso.cv(G=X, Y=Y, E=E, C=U, family="binomial", 
                    grid_min_ratio=0.15, grid_size=20, nfolds=4,
                    type_measure="auc",
                    normalize=TRUE, verbose=TRUE)
stop = Sys.time() - start; stop

# obtaining results
coefs = gesso.coef(fit=gesso_cv$fit, lambda=gesso_cv$lambda_min)
sum(coefs$beta_gxe != 0)
snps_lambda_min = colnames(X)[coefs$beta_gxe != 0]
snps_lambda_min
causalSNPs[6:10]

# summarizing results
gesso_results_gxe = tibble(name=colnames(X)[coefs$beta_gxe != 0],
                           coef=coefs$beta_gxe[coefs$beta_gxe != 0],
                           scaled_coef=coefs$beta_gxe[coefs$beta_gxe != 0] * 
                             apply((X*E)[,coefs$beta_gxe != 0], 2, sd)) %>%
                    arrange(desc(abs(scaled_coef)))
gesso_results_gxe

# alternative way to obtain coefficients
coefs = gesso.coefnum(cv_model=gesso_cv, target_b_gxe_non_zero=6, less_than=TRUE)
sum(coefs$beta_gxe != 0)
snps_num_gxe = colnames(X)[coefs$beta_gxe != 0]
snps_num_gxe
causalSNPs[6:10]
```


# Evaluation of the results with principal selection metrics
```{r}
selection_metrics = function(snp_names, true_interactions) {
  snp_indices = as.numeric(sapply(strsplit(snp_names, split="[.]"), 
                                  function(x) x[2]))
  
  TP_num = length(intersect(snp_indices, true_interactions))
  TN_num = (p - length(true_interactions)) - (length(snp_indices) - TP_num)
  
  precision = TP_num/ length(snp_names)
  sensitivity = TP_num/ length(true_interactions)
  specificity = TN_num / (p - length(true_interactions))
  
  return(list(precision=precision, 
              sensitivity=sensitivity, 
              specificity=specificity))
}

selection_lambda_min = selection_metrics(snps_lambda_min, causalSNPs[6:10])
selection_num_gxe = selection_metrics(snps_num_gxe, causalSNPs[6:10])
cbind(selection_lambda_min, selection_num_gxe)
```

# Single-marker p-vals for the selected SNPs
```{r}
gwis_alpha = 0.05/p; gwis_alpha
for (g_ in snps_lambda_min){
  print(g_)
  g = X[,g_]
  reg = summary(glm(Y ~ g*E + U, family="binomial"))   
  print(round(reg$coefficients[c("g", "g:E"),], 5))
}
```

# Explore individual SNPs 
```{r}
snp = 132
snp_name = paste0("SNP.", snp, "")
cor_ = tibble(cor=as.vector(cor(X[,-snp], X[,snp_name])),
              name=colnames(X[,-snp]))
print(cor_ %>% arrange(desc(cor_)) %>% filter(cor >= 0.3) %>% arrange(name), n=50) 
```